---
title: "Index"
author: "Monica Pittiglio"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
# Just a load of all the packages I used or attempted to use for this project.
knitr::opts_chunk$set(echo = TRUE)
library(datasets)
library(plyr)
library(tidyverse)
library(kableExtra)
library(knitr)
library(readr)
library(lattice)
library(ggplot2)
library(grid)
library(directlabels)
library(tidyquant)
library(dplyr)
library(sf)
library(rgeos)
library(rnaturalearth)
```

# COVID-19 Research Project
### BIOL 390: Reproducable Research, Summer 2020
**Copyright (2020) Monica Pittiglio**

```{r, include = FALSE}
# Load of all coronavirus datasets, dataframes.
Global <- read_csv("~/GitHub/COVID-19/raw_data/WHO-COVID-19-global-data (1).csv")

# Comprehensive COVID data, filtered for countries and separated by healthcare type
NonUniversal <- Global %>%
  filter(Country %in% c("Bangladesh", "Burundi", "Congo", "Ethiopia", "Jordan", "Kenya", "Nigeria", "Paraguay", "United Arab Emirates", "United Republic of Tanzania", "Uganda", "United States of America"))

Universal <- Global %>%
  filter(Country %in% c("Australia", "Bahrain", "Bhutan", "Botswana", "Brazil", "Brunei Darussalam", "Canada", "Cuba", "Denmark", "Finland", "Georgia", "Greece", "Iceland", "Ireland", "Italy", "Kuwait", "Malta", "New Zealand", "Norway", "Oman", "Portugal", "San Marino", "Saudi Arabia", "South Africa", "Spain", "Sri Lanka", "Sweden", "Trinidad and Tobago", "The United Kingdom"))

# Comprehensive COVID data, filtered for countries
AllCOVID <- Global %>%
  filter(Country %in% c("Australia", "Bahrain", "Bhutan", "Botswana", "Brazil", "Brunei Darussalam", "Canada", "Cuba", "Denmark", "Finland", "Georgia", "Greece", "Iceland", "Ireland", "Italy", "Kuwait", "Malta", "New Zealand", "Norway", "Oman", "Portugal", "San Marino", "Saudi Arabia", "South Africa", "Spain", "Sri Lanka", "Sweden", "Trinidad and Tobago", "The United Kingdom", "Bangladesh", "Burundi", "Congo", "Ethiopia", "Jordan", "Kenya", "Nigeria", "Paraguay", "United Arab Emirates", "United Republic of Tanzania", "Uganda", "United States of America"))

# Calculations of total deaths, both by healthcare type and in total
LastDayGlobal <- filter(Global, Date_reported == "2020-08-02")
sum(LastDayGlobal$Cumulative_deaths)

LastDayNonUniversal <- filter(NonUniversal, Date_reported == "2020-08-02")
sum(LastDayNonUniversal$Cumulative_deaths)

LastDayUniversal <- filter(Universal, Date_reported == "2020-08-02")
sum(LastDayUniversal$Cumulative_deaths)
```

```{r, include = FALSE}
# Load population data, filtered by country and separated by healthcare type
Population <- read_csv("~/GitHub/COVID-19/output/Population.csv")

NonUniversalPop <- Population %>%
  filter(Country_name %in% c("Bangladesh", "Burundi", "Congo", "Ethiopia", "Jordan", "Kenya", "Nigeria", "Paraguay", "United Arab Emirates", "United Republic of Tanzania", "Uganda", "United States of America"))

UniversalPop <- Population %>%
  filter(Country_name %in% c("Australia", "Bahrain", "Bhutan", "Botswana", "Brazil", "Brunei Darussalam", "Canada", "Cuba", "Denmark", "Finland", "Georgia", "Greece", "Iceland", "Ireland", "Italy", "Kuwait", "Malta", "New Zealand", "Norway", "Oman", "Portugal", "San Marino", "Saudi Arabia", "South Africa", "Spain", "Sri Lanka", "Sweden", "Trinidad and Tobago", "The United Kingdom"))
```

## Abstract
There is much debate worldwide over which healthcare type is superior - universal or non-universal. With previous data backing the claim that broader health coverage increase access to necessary care, this report set out to determine whether countries with universal healthcare had lower COVID mortality rates than those with non-universal healthcare.

World COVID death data and recent population data were both examined for total deaths and deaths as a proportion of population. The results disproved the hypothesis. While countries with universal healthcare averaged fewer deaths per country, they had more total deaths than countries with non-universal healthcare, despite having almost half as many total citizens. They averaged four times more deaths per capita than countries with non-universal healthcare, and totaled ten times more. This could be due to a decrease in the quality of care provided by practitioners employed under universal healthcare regulations, but the real cause will require further analysis.

## Introduction
Universal government-funded health systems are such a complex beast that only 31 of the world's 32 developed nations have been able to make them work. In countries with non-universal healthcare, up to 90% of the population may be uninsured. There is data that exists already linking broader health coverage to better access to necessary care and improved overall health (Moreno-Serra and Smith, 2012). Thus, it is expect that healthcare type is also indicative of how well-equiped a country is to handle a global pandemic. 

Healthcare type is expected to impact how likely citizens are to seek treatment once they develop COVID symptoms, the prevelance of untreated underlying health conditions, and how healthy the general population is. All of these items will directly impact death numbers. This report examines the COVID-related deaths in countries with universal healthcare, and compares them to the COVID-related deaths in countries with non-universal healthcare.

## Hypothesis
I hypothesize that countries with non-universal health insurance systems experience greater COVID-19 mortality rates than countries with universal government-funded health systems.

In countries that have non-universal health insurance systems, some citizens may have private insurance, some may be eligible for subsidized public healthcare, and some may be completely without insurance. Countries that fall under this category are Bangladesh, Burundi, the Democratic Republic of Congo, Ethiopia, Jordan, Kenya, Nigeria, Paraguay, Tanzania, Uganda, the United Arab Emirates, and of course, the United States. Egypt, India, and Indonesia also fall into this category but will be left out of this report, as their healthcare systems are in transition.

In countries with universal government-funded health systems, healthcare is available to all citizens, no matter what. Countries that fall under this category are Australia, Bahrain, Bhutan, Botswana, Brazil, Brunei, Canada, Cuba, Denmark, Finland, Georgia, Greece, Iceland, Ireland, Italy, Kuwait, Malta, New Zealand, Norway, Oman, Portugal, San Marino, Saudi Arabia, South Africa, Spain, Sri Lanka, Sweden, Trinidad and Tobago, and the United Kingdom. Taiwan and North Korea also fall into this category but will be left out of this report, as they have not released their coronavirus data.

Other types of healthcare do exist. Some governments provide universal public insurance, universal public-private insurance, or universal private insurance. Those healthcare systems, however, are not examined in this report.

```{r, echo = FALSE, include = FALSE}

# Load data that includes population and healthcare type by country
RePopulation_and_Healthcare <- read_csv("~/GitHub/COVID-19/output/Re Population and Healthcare.csv")

NewPopAndHealth <- RePopulation_and_Healthcare %>%
  filter(Country_name %in% c("Australia", "Bahrain", "Bhutan", "Botswana", "Brazil", "Brunei", "Canada", "Cuba", "Denmark", "Finland", "Georgia", "Greece", "Iceland", "Ireland", "Italy", "Kuwait", "Malta", "New Zealand", "Norway", "Oman", "Portugal", "San Marino", "Saudi Arabia", "South Africa", "Spain", "Sri Lanka", "Sweden", "Trinidad and Tobago", "United Kingdom", "Bangladesh", "Burundi", "Congo", "Ethiopia", "Jordan", "Kenya", "Nigeria", "Paraguay", "United Arab Emirates", "Tanzania", "Uganda", "United States"))

# Creation of data set to be used in map below.
World <- ne_countries(scale = "medium", returnclass = "sf")
class(NewPopAndHealth)

Map <- full_join(World, NewPopAndHealth, by = c("name" = "Country_name"))
```

```{r, echo = FALSE, fig.width = 14, fig.align = "center"}
# Map of countries examined in this report, by healthcare type.
ggplot(data = Map) +
    geom_sf(aes(fill = Healthcare)) +
  scale_fill_manual(values = c("red", "blue")) +
   ggtitle("Figure 1: Countries Examined in This Report, by Healthcare Type")
```

## Analysis
First, individual healthcare types were examined by viewing the cumulative deaths per country, grouped by healthcare type. Keep in mind that the Y axis of the following two graphs are on log scales, making the outliers appear less distinguished but the rest of the data more differentiable.

```{r, fig.align = "center", fig.height = 9, echo = FALSE, message = FALSE, warning = FALSE}
# Line graph of total deaths by country, non-universal healthcare.
# I made this graph abnormally tall so that it wasn't too squished!!!
ggplot(NonUniversal, aes(x = Date_reported, y = Cumulative_deaths, color = Country)) +
  geom_line() +
  scale_y_log10() +
   ggtitle("Figure 2: Cumulative Deaths Over Time in Countries with Non-Universal Healthcare") +
  labs(x = "Month", y = "Cumulative Deaths") +
  geom_dl(aes(label = Country), method = list(dl.combine("last.points")), cex = 0.8) +
  coord_x_date(xlim = c("2020-01-15", "2020-11-01")) +
  theme(legend.position = "none")
```

```{r, echo = FALSE}
# Table of statistics for deaths, non-universal healthcare. Compliments above plot.
knitr::kable(LastDayNonUniversal %>% summarize(N = n(), 
              Sum = sum(Cumulative_deaths),
			 			  Mean = mean(Cumulative_deaths), 
			 			  Median = median(Cumulative_deaths), 
			 			  StdDev = sd(Cumulative_deaths)), 
			 align = c('l','c','c','c','c'), 
			 digits = 2,
			 format = "html", 
			 caption = "**Table 1**: Summary of Deaths in Countries with Non-Universal Healthcare",
			 table.attr = "style='width:70%;'") %>% kable_styling(position = "center")
```

The total number of deaths as of August 2nd in countries with non-universal healthcare is 157,786. The United States of America leads the pack with a cumulative 152,630 deaths. This is followed by Bangladesh, which has a cumulative 3,154 deaths, and Nigeria, which has a cumulative 883 deaths. These top three countries account for 99% of the deaths in countries that have non-universal healthcare. The United States of America alone comprises 97% of the total.

```{r, fig.align = "center", fig.height = 12, echo = FALSE, message = FALSE, warning = FALSE}
# Line graph of total deaths by country, universal healthcare.
# I made this graph abnormally tall so that it wasn't too squished!!!
ggplot(Universal, aes(x = Date_reported, y = Cumulative_deaths, color = Country)) +
  geom_line() +
  scale_y_log10() +
  ggtitle("Figure 3: Cumulative Deaths Over Time in Countries with Universal Healthcare") +
  labs(x = "Month", y = "Cumulative Deaths") +
  geom_dl(aes(label = Country), method = list(dl.combine("last.points")), cex = 0.8) +
  coord_x_date(xlim = c("2020-01-15", "2020-10-01")) +
  theme(legend.position = "none")
```

```{r, echo = FALSE}
# Table of statistics for deaths, universal healthcare. Compliments above plot.
knitr::kable(LastDayUniversal %>% summarize(N = n(),
              Sum = sum(Cumulative_deaths),
			 			  Mean = mean(Cumulative_deaths), 
			 			  Median = median(Cumulative_deaths), 
			 			  StdDev = sd(Cumulative_deaths)), 
			 align = c('l','c','c','c','c'), 
			 digits = 2,
			 format = "html", 
			 caption = "**Table 2**: Summary of Deaths in Countries with Universal Healthcare",
			 table.attr = "style='width:70%;'") %>% kable_styling(position = "center")
```

The total number of deaths as of August 2nd in countries with universal healthcare is 234,315. Brazil leads the pack with a cumulative 92,475 deaths. This is followed by the United Kingdom, which has a cumulative 46,193 deaths, and Italy, which has a cumulative 35,146 deaths. These top three countries account for 74% of the deaths in countries that have universal healthcare.

Upon comparison, we can see that countries with non-universal healthcare average more deaths than those with universal healthcare. However, they have an enormous standard deviation, no doubt due to the pull of the United States' death toll, which is 48 times that of the next country in the dataset. Countries with universal healthcare had the larger sum and median, with a standard deviation 45% smaller than that of the non-universal healthcare countries. In general, countries with universal healthcare tend towards more coronavirus deaths than those with non-universal healthcare.

To quote Jonathan Swan in a recent interview with President Trump, "It's surely a relevant statistic to say if the US has 'x' population and 'x' percentage of death of that population..." He is correct - even if the president doesn't want us to, we must look at deaths per capita in order to compare the datasets properly. To begin the process, we'll examine population count separated by healthcare type.

```{r, echo = FALSE, include = FALSE}
# Load data that includes population and healthcare type by country.
Population_and_Healthcare <- read_csv("~/GitHub/COVID-19/output/Population and Healthcare.csv")

AllPopAndHealth <- Population_and_Healthcare %>%
  filter(Country_name %in% c("Australia", "Bahrain", "Bhutan", "Botswana", "Brazil", "Brunei Darussalam", "Canada", "Cuba", "Denmark", "Finland", "Georgia", "Greece", "Iceland", "Ireland", "Italy", "Kuwait", "Malta", "New Zealand", "Norway", "Oman", "Portugal", "San Marino", "Saudi Arabia", "South Africa", "Spain", "Sri Lanka", "Sweden", "Trinidad and Tobago", "The United Kingdom", "Bangladesh", "Burundi", "Congo", "Ethiopia", "Jordan", "Kenya", "Nigeria", "Paraguay", "United Arab Emirates", "United Republic of Tanzania", "Uganda", "United States of America"))
```

```{r, fig.align = "center", fig.width = 14, fig.height = 7, echo = FALSE}
# Point plot with lines of 2019 population, by healthcare type.
AllPopAndHealth%>%
  mutate(CN = fct_reorder(Country_name, Year_2019)) %>%
  ggplot( aes(x = Year_2019, y = CN, color = Healthcare)) +
  geom_point(size = 4) +
  geom_segment(aes(xend = 0, yend = Country_name), size = 2) +
  ggtitle("Figure 4: Total population in Countries with Both Healthcare Types") +
  labs(x = "Population", y = "Country")
```

```{r, echo = FALSE}
# Table of statistics for population counts by healthcare type. Compliments above plot.
knitr::kable(AllPopAndHealth %>%
              group_by(Healthcare) %>%
              summarize(N = n(),
              Sum = sum(Year_2019),
			 			  Mean = mean(Year_2019), 
			 			  Median = median(Year_2019), 
			 			  StdDev = sd(Year_2019), .groups = 'drop'), 
			 align = c('l','c','c','c','c'), 
			 digits = 2,
			 format = "html", 
			 caption = "**Table 3**: Population Summary by Healthcare Type",
			 table.attr = "style='width:70%;'") %>% kable_styling(position = "center")
```

This graph's axis is not on a log scale, so we can clearly see that even though there are less countries with non-universal healthcare, they have a larger summed population than countries that have universal healthcare. In fact, the four most highly-populated countries that have non-universal healthcare account for 51% of the total population between all 41 countries in this graph. The average population of the non-universal healthcare countries is also 4 times larger than the average population of the universal healthcare countries. That said, their standard deviation is also much larger, no doubt due to the United States' massive population.

```{r echo = FALSE}
# Joined population and healthcare data, filtered for usable data.
All <- inner_join(AllPopAndHealth, AllCOVID, by = c("Country_name" = "Country"))

# Created per capita column
AllByPop <- All %>%
  filter(Date_reported == "2020-08-02") %>%
  filter(Country_name %in% c("Australia", "Bahrain", "Bhutan", "Botswana", "Brazil", "Brunei Darussalam", "Canada", "Cuba", "Denmark", "Finland", "Georgia", "Greece", "Iceland", "Ireland", "Italy", "Kuwait", "Malta", "New Zealand", "Norway", "Oman", "Portugal", "San Marino", "Saudi Arabia", "South Africa", "Spain", "Sri Lanka", "Sweden", "Trinidad and Tobago", "The United Kingdom", "Bangladesh", "Burundi", "Congo", "Ethiopia", "Jordan", "Kenya", "Nigeria", "Paraguay", "United Arab Emirates", "United Republic of Tanzania", "Uganda", "United States of America")) %>%
  mutate(DeathsPerCapita = Cumulative_deaths / Year_2019) %>%
  mutate(CasesPerCapita = Cumulative_cases / Year_2019)

UniversalByPop <- All %>%
  filter(Country_name %in% c("Australia", "Bahrain", "Bhutan", "Botswana", "Brazil", "Brunei Darussalam", "Canada", "Cuba", "Denmark", "Finland", "Georgia", "Greece", "Iceland", "Ireland", "Italy", "Kuwait", "Malta", "New Zealand", "Norway", "Oman", "Portugal", "San Marino", "Saudi Arabia", "South Africa", "Spain", "Sri Lanka", "Sweden", "Trinidad and Tobago", "The United Kingdom")) %>%
  mutate(DeathsPerCapita = Cumulative_deaths / Year_2019) %>%
  mutate(CasesPerCapita = Cumulative_cases / Year_2019) %>%
  mutate(NewDeathsPerCapita = New_deaths / Year_2019) %>%
  mutate(NewCasesPerCapita = New_cases / Year_2019)

NonUniversalByPop <- All %>%
  filter(Country_name %in% c("Bangladesh", "Burundi", "Congo", "Ethiopia", "Jordan", "Kenya", "Nigeria", "Paraguay", "United Arab Emirates", "United Republic of Tanzania", "Uganda", "United States of America")) %>%
  mutate(DeathsPerCapita = Cumulative_deaths / Year_2019) %>%
  mutate(CasesPerCapita = Cumulative_cases / Year_2019) %>%
  mutate(NewDeathsPerCapita = New_deaths / Year_2019) %>%
  mutate(NewCasesPerCapita = New_cases / Year_2019)

LastUniversalByPop <- UniversalByPop %>%
  filter(Date_reported == "2020-08-02")

LastNonUniversalByPop <- NonUniversalByPop %>%
  filter(Date_reported == "2020-08-02")
```

```{r, fig.align = "center", fig.width = 14, fig.height = 6, echo = FALSE}
# Point plot with lines of deaths per capita
AllByPop%>%
  mutate(PC = fct_reorder(Country_name, DeathsPerCapita)) %>%
  ggplot( aes(x = DeathsPerCapita, y = PC, color = Healthcare)) +
  geom_point(size = 4) +
  geom_segment(aes(xend = 0, yend = Country_name), size = 2) +
  ggtitle("Figure 5: Total Deaths Per Capita in Countries with Both Healthcare Types") +
  labs(x = "Population", y = "Country")
```

```{r, echo = FALSE}
# Table of statistics for population counts by healthcare type. Compliments above plot.
knitr::kable(AllByPop %>%
              group_by(Healthcare) %>%
              summarize(N = n(),
              Sum = sum(DeathsPerCapita),
			 			  Mean = mean(DeathsPerCapita), 
			 			  Median = median(DeathsPerCapita), 
			 			  StdDev = sd(DeathsPerCapita), .groups = 'drop'), 
			 align = c('l','c','c','c','c'), 
			 digits = 6,
			 format = "html", 
			 caption = "**Table 4**: Summary of Deaths per Capita by Healthcare Type",
			 table.attr = "style='width:70%;'") %>% kable_styling(position = "center")
```

Once we compared deaths as a proportion of population, the graph completely flipped. While countries with universal healthcare have smaller populations, they tend to have far more deaths per capita than countries with non-universal healthcare - a total of 10 times more, actually, and an average of 4 times more. Seven of the non-universal healthcare countries are among the bottom ten deaths per capita, and 18 of the universal healthcare countries are among the top 20 deaths per capita. When you consider that there are only 41 countries analyzed in this report, these results are remarkably clear. Notably, San Marino had the smallest population but the most deaths per capita.

For a more in-depth look at what's going on in each country, we can view individual graphs of daily deaths per capita in each country. Keep in mind that the y-axis are on log 10 scales for viewing ease.

```{r, fig.align = "center", fig.width = 10, fig.height = 7, echo = FALSE, message = FALSE, warning = FALSE}
# Individual graphs for every country with non-universal healthcare, of daily deaths per capita.
# I want to stress that I spent roughly 5 hours trying to figure out how to add regression lines to facited plots, and while it is technically possible, it requires a significant formatting change because you have to add it to each graph individually. The resulting graphs look horrible when there are this many plots.
ggplot() +
  geom_line(data = NonUniversalByPop, aes(x = Date_reported, y = NewDeathsPerCapita, color = "Deaths")) +
  facet_wrap(~ Country_name) +
  ggtitle("Figure 6: Daily Deaths Per Capita in Countries with Non-Universal Healthcare") +
  labs(x = "Month", y = "Daily Deaths Per Capita") +
  scale_y_log10() +
  theme(legend.position = "none")
```

```{r, echo = FALSE}
# Joins logical healthcare type data with the rest of the data analyzed in above plot
NonUniversalFinal <- right_join(NewPopAndHealth, NonUniversalByPop, by = "Country_name")

# ANOVA table of new deaths per capita compared to healthcare type. Compliments Figure 6.
knitr::kable(anova(lm(NewDeathsPerCapita ~ UHC, data = NonUniversalFinal)),
             digits = 15,
             align = c('l','c','c','c','c'),
			 format = "html",
			 caption = "**Table 5**: ANOVA of New Deaths per Capita in Countries with Non-Universal Healthcare",
			 table.attr = "style='width:70%;'") %>% 
  kable_styling(position = "center")
```

Burundi and the United Republic of Tanzania have done a great job lately of getting their deaths under control. Paraguay, Congo, and the United Arab Emirates appear to have a mixture of good and bad days, where deaths per capita can range greatly. The United States of America, Bangladesh, and Ethiopia appear to be losing the fight to COVID - their deaths per capita are large and continuting to grow.

```{r, fig.align = "center", fig.width = 10, fig.height = 7, echo = FALSE, message = FALSE, warning = FALSE}
# Individual graphs for every country with universal healthcare, of daily deaths per capita.
ggplot() +
  geom_line(data = UniversalByPop, aes(x = Date_reported, y = NewDeathsPerCapita, color = "Cases")) +
  facet_wrap(~ Country_name) +
  ggtitle("Figure 7: Daily Deaths Per Capita in Countries with Universal Healthcare") +
  labs(x = "Month", y = "Daily Deaths Per Capita") +
  scale_y_log10() +
  theme(legend.position = "none")
```

```{r, echo = FALSE}
# Joins logical healthcare type data with the rest of the data analyzed in above plot
UniversalFinal <- right_join(NewPopAndHealth, UniversalByPop, by = "Country_name")

# ANOVA table of new deaths per capita compared to healthcare type. Compliments Figure 7.
knitr::kable(anova(lm(NewDeathsPerCapita ~ UHC, data = UniversalFinal)),
             digits = 15,
             align = c('l','c','c','c','c'),
			 format = "html",
			 caption = "**Table 6**: ANOVA of New Deaths per Capita in Countries with Universal Healthcare",
			 table.attr = "style='width:70%;'") %>% 
  kable_styling(position = "center")
```

Bhutan is the real superstar of this report, with zero deaths. Italy, the United Kingdom, Spain, Ireland, and Canada appear to be doing a great job of drastically decreasing their daily deaths per capita, as clear trends are visible. The same can not be said of Brazil and South Africa, whose daily cases are still on the rise.

The countries with universal healthcare have a sum of squares 3000 times larger than those with universal healthcare, indicating more variability in the number of new cases every day. Some countries with universal healthcare may be handling the virus well, but their efforts are offset globally by those who are not.

## Results and discussion

While countries with non-universal healthcare average more deaths than those with universal healthcare, this is due almost entirely to the pull from one country. Those with universal healthcare had a larger sum and median of total deaths - an expected outcome, as there are 29 countries with universal healthcare and only 12 with non-universal.

Once population gets brought into consideration, the results are unexpected. Countries with non-universal healthcare had both larger total and larger average populations than those with universal healthcare, despite being comprised of far fewer counties. One-quarter of the countries with non-universal healthcare (n = 4) accounted for over 50% of the total population in all 41 countries that we looked at. It's certainly possible that it's more complicated, expensive, or difficult to implement universal healthcare in countries with large populations - private insurance shifts the burden off the the country's government.

Countries with universal healthcare had more deaths per capita despite having almost half as many total citizens. They averaged four times more deaths per capita than countries with non-universal healthcare, and totaled ten times more. This could be because universal healthcare makes for long wait times to be seen by doctors, and may cause doctors to cut corners due to monetary issues. There is also much debate as to whether public, government-funded hospitals function at the same standards and efficiency as private hospitals.

The countries with universal healthcare also saw more variability in the number of new cases every day. While a few countries with universal healthcare are seeing a trending decrease in new cases, there are many more that are seeing an increase in cases.

## Conclusions
The data does not support the hypothesis. While countries with non-universal healthcare average more deaths overall than countries with non-universal healthcare, that is due almost entirely to the data from one country - the United States. Countries with universal healthcare saw more deaths, both in total and per capita. More research will be needed to figure out why this is the case.

## Viewing this project
The completed project files can be accessed online at https://MonicaPittiglio.github.io/COVID-19/. The repo for this reproducible project is also publicly available at https://github.com/MonicaPittiglio/COVID-19.

## Refrences
Moreno-Serra R, Smith P. Does progress towards universal health coverage improve population health? *The Lancet.* 2012;
*380*:917-23. doi:10.1016/S0140-6736(12)61039-3.


“Population, Total.” World Bank, World Bank Group, 2020, data.worldbank.org/indicator/SP.POP.TOTL.


WHO Coronavirus Disease (COVID-19) Dashboard. World Health Organization, 2020, covid19.who.int/.