---
title: "Index"
author: "Monica Pittiglio"
date: "7/22/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(kableExtra)
library(readr)
library(lattice)
```

# COVID-19 Research Project
### BIOL 390: Reproducable Research, Summer 2020
**Copyright (2020) Monica Pittiglio**
```{r, include = FALSE}
Global <- read_csv("~/GitHub/COVID-19/raw_data/WHO-COVID-19-global-data.csv")

NonUniversal <- Global %>%
  filter(Country %in% c("Bangladesh", "Burundi", "Congo", "Ethiopia", "Jordan", "Kenya", "Nigeria", "Paraguay", "United Arab Emirates", "Tanzania", "Uganda", "United States of America"))

Universal <- Global %>%
  filter(Country %in% c("Australia", "Bahrain", "Bhutan", "Botswana", "Brazil", "Brunei", "Canada", "Cuba", "Denmark", "Finland", "Georgia", "Greece", "Iceland", "Ireland", "Italy", "Kuwait", "Malta", "New Zealand", "North Korea", "Norway", "Oman", "Portugal", "San Marino", "Saudi Arabia", "South Africa", "Spain", "Sri Lanka", "Sweden", "Taiwan", "Trinidad and Tobago", "United Kingdom"))
```

## Abstract (tl;dr)

## Introduction
Universal government-funded health systems are such a complex beast that only 31 of the world's 32 developed nations have been able to make them work.

## Hypothesis
I hypothesize that countries with non-universal health insurance systems experience greater COVID-19 mortality rates than countries with universal government-funded health systems.

In countries that have non-universal health insurance systems, some citizens may have private insurance, some may be eligible for subsidized public healthcare, and some may be completely without insurance. Countries that fall under this category are Bangladesh, Burundi, the Democratic Republic of Congo, Ethiopia, Jordan, Kenya, Nigeria, Paraguay, the United Arab Emirates, Tanzania, Uganda, and of course, the United States. Egypt, India, and Indonesia also fall into this category but will be left out of this report, as their healthcare systems are in transition. n = 12.

In countries with universal government-funded health systems, healthcare is available to all citizens, no matter what. Countries that fall under this category are Australia, Bahrain, Bhutan, Botswana, Brazil, Brunei, Canada, Cuba, Denmark, Finland, Georgia, Greece, Iceland, Ireland, Italy, Kuwait, Malta, New Zealand, North Korea, Norway, Oman, Portugal, San Marino, Saudi Arabia, South Africa, Spain, Sri Lanka, Sweden, Taiwan, Trinidad and Tobago, and the United Kingdom. n = 31.

## Datasets used
My main dataset is the World Health Organizaion's international dataset, which includes number of cases and deaths reported both daily and cumulatively, by country. This dataset is updated daily and can be found at https://covid19.who.int/.

```{r, echo = FALSE}
ggplot(NonUniversal, aes(x = Date_reported, y = Cumulative_cases)) +
  geom_line() +
  facet_wrap(~ Country)

ggplot(Universal, aes(x = Date_reported, y = Cumulative_cases)) +
  geom_line() +
  facet_wrap(~ Country)
```

## Description of analyses

## Conclusions

## Viewing this project
The completed project files can be accessed online at file:///C:/Users/Owner/Documents/GitHub/COVID-19/README.html. The repo for this reproducible project is also publicly available at https://github.com/MonicaPittiglio/COVID-19.
